---
title: "Titanic struggles with R"
output: html_notebook
---

This [R Markdown](http://rmarkdown.rstudio.com) notebook is an exploration of
the data for passengers from the Titanic disaster.

We first read the dataset containing information about the passengers (not
crew) who were on the RMS Titanic when it sank.

See the [Titanic dataset
page](https://github.com/odsti/datasets/tree/master/titanic) for more detail
on this dataset.

You might also want to look at [Encylopedia
Titanica](https://www.encyclopedia-titanica.org/titanic-statistics.html) for
extensive background on the ship, its passengers and crew.

```{r}
# Read the dataset directly from the file.
# After this, the variable "titanic" is a data frame.
titanic <- read.csv("titanic_clean.csv")
```

Here and throughout the notebook, the exercise chunks have `...` markers.
These are blanks for you to fill in with the correct code or variables.

Your first task is to show the names of the columns in the data frame:

```{r}
#- Use the "names" function to show the names of the columns in "titanic"
#- Replace the ... with your code.
names(...)
```

Show the number of rows (passengers) and columns (variables), using the `dim`
command:

```{r}
#- Use "dim" to show the number of rows and columns of the "titanic" data
#- frame.
dim(...)
```

Show the first 6 passengers (rows) in the data frame using the `head` function:

```{r}
#- Show the top 6 passengers with the "head" function.
...
```

As you get older, you get richer (at least, that's how it seems to us, your
instructors).  So, are first class passengers generally older than third class
passengers?

Do a box plot of the ages of the passengers as a function of class.  Have
a look to see if the first class passengers tend to be older.

Hint - make sure you've got all three passenger classes on your x axis.  You'll
probably need the `~` syntax that you saw in the tutorial. Remember, you can
read the `~` operator as "as a function of".

```{r}
#- Do a box plot of the distribution of ages as a function of passenger class.
#- Replace the ... with your code.
boxplot(... ~ ...)
```

I am sure we all believe that female passengers were more likely to be put onto
the lifeboats. Now we will check if that's really true.

Use the `table` function to show the counts of male and female passengers
overall.  If you can't remember the column names, have a look at the result of
the `names(titanic)` chunk above, or the column names in the View tab.

```{r}
#- Use "table" to show the counts of male and female passengers.
table(...)
```

Now use the `survived` column to give a table with a breakdown of how many
male and female passengers survived.  Show the table.  Check the tutorial for
how to break down counts by more than one column.

```{r}
#- Make a new table called "mf_survived" with the numbers of men and women who
#- survived.
mf_survived <- table(...)
# Show the table.
mf_survived
```

Show these counts graphically as proportions using `mosaicplot`. Check the
help on `mosaicplot` for the input you need.  See whether you agree that women
were more likely to be saved than men.

```{r}
#- Make a graphical plot of the table of male / female survival numbers with
#- "mosaicplot".
mosaicplot(...)
```

We also have a strong suspicion that the first class passengers had a better
chance of being saved than the second or third class passengers.  Let's check
if that is true.

First make a table breaking down the counts in each passenger class by whether
they survived.  Show that table in the notebook.

```{r}
#- Make a new table called "pclass_survived" with the numbers of first, second,
#- third class passengers who survived.
pclass_survived <- table(...)
# Show the table.
pclass_survived
```

Show the proportions who survived by passenger class graphically using
`mosaicplot`.

```{r}
#- Show the table of survival numbers graphically with "mosaicplot"
mosaicplot(pclass_survived)
```

We also think that children had a better chance of being saved.  We guess that
children in first class are a bit more likely to be saved than children in
third class.  To look at this we first use the `subset` command to make a new
data frame that only includes the passengers with ages less than 15.  Check
back at the tutorial for how to select a subset of a data frame according to
a condition like "age less than 15".  Show the first 6 cases of this new data
frame to check you do in fact have the passengers younger than 15.

```{r}
#- Make a new data frame called "children" containing only the passengers
#- younger than 15.  Use the subset command.
children <- subset(...)
# Use "head" to check that "children" only has passengers younger than 15.
head(children)
```

Use your new `children` data frame to make a table breaking down the counts of
children in the different passenger classes, by whether or not they survived.

Show the table.

```{r}
#- Use the "children" data frame to make a table called "class_children"
#- giving counts of children for each passenger class, broken down by whether
#- or not they survived.
class_children <- ...
# Show the table.
class_children
```

Show the `class_children` table you've just made with `mosaicplot`.  Are
higher-class children more likely to survive?

```{r}
#- Use "mosaicplot" to show the proportions of children of each class who
#- survived.
...
```

Finally, we might wonder if the crew of the Titanic were national chauvinists.
Did they, for example, prefer to save the passengers from good old
Southampton, over those from Ireland (Queenstown) or France (Cherbourg)?  Did
the passengers' survival also depend on where they embarked?

We already know that the chances of being saved were better if you were a
female, and in the higher ticket classes.  To control for that, let us first
select the women in third class.  Use the `subset` function to select the
passengers that are `3rd` class AND `female`.  Refer back to the tutorial for
how to combine conditions like this using the subset function.  Remember that
both `3rd` and `female` are bits of *text*, so you will need to use quotes so R
knows these are bits of text, like this: `"female"`, `"3rd"`.


```{r}
#- Use "subset" to make a new data frame "third_class_female" that contains all
#- the passengers that are both female and in third class.
#- Check the tutorial for examples of selecting rows on more than one
#- criterion.
third_class_female <- subset(...)
# Show the first 6 rows of this new data frame.
head(third_class_female)
```

Make a table that breaks down counts of new `third_class_female` data frame by
where they embarked, and whether they survived. Show this table.  Have a look at the top of the notebook to see which column you need.

```{r}
#- Use the "third_class_female" data frame to make a table called
#- "port_by_saved" giving counts of third class female passengers for each port
#- they embarked, broken down by whether or not they survived.
port_by_saved <- ...
# Show the table
port_by_saved
```

Show the proportions in this table with "mosaicplot".  Did the crew prefer to
save English women?

```{r}
#- Show the new "port_by_saved" table with a mosaicplot
...
```

That's it.

When you have finished, make sure that all the chunks execute correctly.  To
do that, go to the Run button at the top of the notebook, and click "Restart R
and Clear Output".  Then go to the top of the notebook, and run each chunk in
turn, to make sure that they run without error, and give you the output you
expect.